Stabilizing selection, archaic introgression, and the distribution of shared functional variation

### 2023-06-27 doc/manuscript.tex
outline some results and discussion

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/neanderthal_admixture.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:neand-to-human}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/human_admixture.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:human-to-neand}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/linkage_predictions.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:linkage-pred}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics{../figures/linkage_simulation.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:linkage-sim}
\end{figure}

\clearpage

@@

\begin{figure}[ht!]
    \centering
    %\includegraphics{}
    \caption{
        \textbf{Dynamics of additive genetic variance in two populations with admixture.}
        Coupled with the equations for the decay of $V_G$ due to drift and selection,
        and the increase in $V_G$ through mutation, show that the expected dynamics
        are matched by simulations without linkage.
    }
    \label{fig:VG_dynamics}
\end{figure}

@@

\subsection{The role of linkage}

\section{Discussion}

\section{Methods}

>>>>

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/neanderthal_admixture.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:neand-to-human}
\end{figure}

\begin{itemize}
    \item In the demographic model more closely resembling inferred history
        connecting humans and Neanderthals (Figure~\ref{fig:neand-to-human}A),
        we see that demographic history and the distribution of effect sizes
        of new mutations impact $V_G$
    \item Additive genetic variance increases after introgression in both directions,
        which is expected to result in a period of decreased mean fitness
    \item Population contractions decrease $V_G$, as the increased rate of drift
        reduces diversity at loci underlying the trait
    \item TODO: compare to linked simulations
\end{itemize}

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/human_admixture.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:human-to-neand}
\end{figure}

\begin{itemize}
    \item Because we know if a mutation was previously segregating in the population
        receiving migrants, we can partition variation by contributions from
        segregating vs introgressed variants (whether introduced derived mutations
        or reintroduced ancestral mutations)
    \item At low introgression proportions, as here, the majority of $V_G$ is still
        contributed by previously segregating mutations, which is then expected to
        be steadily replaced by new mutations (Figure 2). $V_G$ due to introgressed
        variation decays monotonically over time (Figure~\ref{fig:neand-to-human}C,D)
    \item While the total $V_G$ due to introgressed variants decays, we can also
        consider the average contribution of introgressed vs non-introgressed SNPs
        to $V_G$ ($h^2$ per SNP) -- at least for this demographic model and
        DES, the contribution per SNP of introgressed variants is initially lower
        than those of previously segregating variants. These contributions change
        over time, in a way that depends on both the genomic architecture underlying
        the trait and demography
    \item This dependence on both architecture and demography is apparent when
        comparing to $V_G$ contributed by introgressed and non-introgressed variants
        in the early human-to-Neanderthal admixture (Figure~\ref{fig:human-to-neand})
        -- here, again the contribution
        from introgressed variants to $V_G$ decays monotonically, but depending on
        the distribution of effect sizes, relative per-SNP $h^2$ changes over time
        (though in these cases introgressed $h^2$ per SNP remains lower than
        non-introgressed, when considering mutations at matched allele frequencies)
\end{itemize}

\subsection*{The effects of linkage}

\begin{itemize}
    \item So far, we have considered model and simulation without linkage -- in this
        case, the underdominance model provides an excellent approximation
    \item Including linkage between sites dramatically complicates things, making
        theorical or analytical progress difficult
    \item To explore the effects of linkage, we used individual-based simulations
        with a single large chromosome of size 1 Morgan, and varried the total
        mutation rate per haploid and the variance of effect sizes of new mutations
    \item We'll highlight two effects of linkage
        \begin{itemize}
            \item Deviations from expected $V_G$, as well as more complicated
                short-term dynamics after admixture
            \item The reduction in introgressed ancestry in the regions surrounding
                trait-affecting alleles
        \end{itemize}
\end{itemize}

\subsubsection*{Effects of linkage on expected genetic variance}

\begin{figure}[t!]
    \centering
    %\includegraphics{}
    \caption{}
    \label{fig:linkage-deviations}
\end{figure}

\begin{itemize}
    \item Stabilizing selection reduces $V_G$ by selecting against minor alleles
        (underdominance), but with linkage it can also reduce $V_G$ by selecting
        for haplotypes that carry opposite effect alleles that cancel out -- this
        creates LD between positive and negative effect alleles at linked loci,
        known as the Bulmer effect \citep{}
    \item Thus, $V_G$ in simulations with linkage deviate from predictions
        assuming underdominance alone
    \item The severity of this deviation depends on a number of factors: mean
        recombination rate between trait-affecting loci, the total mutation rate,
        the effect sizes (and distribution of effect sizes of new mutations)
    \item In general, lower mutation rates imply lower total $V_G$, lower polygenicity,
        and therefore higher recombination rates between trait-affecting alleles --
        we therefore expect this deviation to be exacerbated with higher mutation
        rates (all else being equal)
\end{itemize}

\subsubsection*{Reduction of introgressed ancestry surrounding an introgressed trait-affecting allele}

\begin{itemize}
    \item Because introgressed trait-affecting alleles are selected against as the
        minor allele, linkage will also remove introgressed ancestry segments in the
        regions surrounding the selected alleles
    \item The expected reduction in introgressed ancestry proportions surrounding
        functional alleles will depend on the effect size of the trait-affecting allele
        and the probability of recombination surrounding that site
    \item We can consider first a deterministic model of the dynamics of introgressed
        allele frequencies at and around an introgressed functional allele -- this
        model considers a single trait-affecting locus, ignoring any effects of
        interference between selected alleles
        \begin{itemize}
            \item Consider a site with fixed differences between the two populations
                (the derived allele is fixed in one, lost in the other)
            \item At the functional site itself, with admixture proportion $f$ from the
                minor parental ancestry, the initial frequency of the derived allele
                is $p_0=f$ or $p_0 = 1-f$, if the derived allele was fixed in the 
                major parental ancestry population
            \item In one generation, the expected allele frequency at the selected locus
                is \[p_{t+1} \approx p_t - s p_t(1-p_t)(1-2p_t),\] where $s=a^2/2V_S$
            \item As discussed earlier, if $p_0<1/2$, $p_t\rightarrow0$, and if $p_0>1/2$,
                $p_t\rightarrow1$ as $t\rightarrow\infty$
            \item Consider a neutral locus separated from the selected locus by
                recombination probability $r$. Initially, the expected frequency
                of introgressed ancestry at that locus is $q_0=f$
            \item The linked ancestry frequency will change due to selection acting on
                the trait-affecting locus, depending on the strength of selection
                at that locus and the level of linkage disequilibrium between
                the two loci
            \item $D=Cov(p,q)$ is the standard covariance measure of LD between loci,
                and $q$ is expected to change as \[q_{t+1} \approx q_t - s D_t(1-2p_t)\]
            \item $D$ also changes over time, due to both selection and recombination,
                so that \[D_{t+1} \approx D_t - r D_t - s D_t (1-2p_t)^2\]
            \item Initially, $D_0 = \pm f(1-f)$, with $D$ being positive if $p_0=f$
                and negative if $p_0=1-f$
            \item Together, this forms a nonlinear system of equations for the
                deterministic change in allele frequencies at the two loci (one
                selected, one neutral) and LD between them -- note that this is
                deterministic in the infinite population size limit
        \end{itemize}
    \item Using this model, we can predict how allele frequencies and LD are expected
        to change over time after admixture, for a given effect size $a$ and 
        recombination rate $r$ (Figure~\ref{fig:linkage-pred}A,B)
    \item As expected, lower effect sizes result in a slower reduction in introgressed
        ancestry frequency, while higher recombination rates more quickly decouple
        the linked ancestry from the selected allele
    \item Thus, the expected reduction in introgressed ancestry is largest for
        larger effect sizes (Figure~\ref{fig:linkage-pred}C), and LD between the
        selected and linked neutral alleles is largest for neutral sites closest
        to the selected allele (Figure~\ref{fig:linkage-pred}D)
\end{itemize}

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/linkage_predictions.pdf}
    \caption{
        \textbf{}
        \aprcomment{Linked allele frequency should always be minor, and labeled
        introgressed ancestry frequency}
    }
    \label{fig:linkage-pred}
\end{figure}

\begin{itemize}
    \item To explore the accuracy of this two-locus model, we compared to
        simulations using the demographic model shown in
        Figure~\ref{fig:toy-admixture}A with introgression fraction $f=0.05$.
        We simulated a single 1 Morgan
        chromosome, with all mutations having effect sizes $\pm a$ ($a=0.02$ or
        $0.05$, with equal probability of being trait-increasing or decreasing)
        and varying the total mutation rate ($\mu=0.001$, $0.0025$, and $0.01$)
    \item For each mutation that was fixed in one population and absent in the
        other (either fixed in the minor or major ancestry population), we
        determined the average introgressed ancestry surrounding those introgressed
        alleles (Figure~\ref{fig:linkage-sim})
    \item The deterministic model provides a very good approximation for linked
        introgressed ancestry when the mutation rate is small
    \item As the mutation rate increases, polygenicity increases so that
        trait-affecting alleles are more densely distributed along the chromosome --
        deviations from the deterministic model are due to interference between
        tightly linked trait-affecting alleles. For the highest mutation rate shown
        here, there will be many trait-affecting alleles within the 1 cM window
        around any focal SNP
\end{itemize}

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/linkage_simulation.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:linkage-sim}
\end{figure}

\section*{Discussion}

\begin{itemize}
    \item Stabilizing selection with a constant optimum is an appropriate
        null model for the dynamics of trait variation and selection on
        alleles underlying those traits
    \item Stabilizing selection results in population genetic predictions
        for patterns of introgressed variation in and around functional
        regions that differ from both load- and incompatibility-based arguments
        \begin{itemize}
            \item Load-based models predict that haplotypes from the less ``loaded''
                population will be favored under introgression in either direction
                -- thus, regions of depleted ancestry after introgression in one
                direction will not be expected to also be introgressed ancestry
                deserts after admixture in the opposite direction
            \item \citet{mueller1942} originally pointed out that hybrid
                incompatibilities should not form at single loci -- classical
                incompatibilities require two or more loci, which are not
                expected to be tightly linked to one another. Thus, under a
                model of hybrid incompatibilities, the minor parent ancestry
                is again selected against, but those loci will be in different
                regions. Again, deserts are therefore not expected to overlap
                if selection is largely due to incompatibilities
            \item Instead, stabilizing selection causes introgressed DNA to be
                depleted at the same loci under bi-directional gene flow
                (could be thought of as single-locus incompatibility effects --
                but these are not incompatibilities in the classical sense,
                as the same effects occur even without admixture)
            \item A prediction we could make is that, if current models are correct
                that there has been introgression in both directions between
                \emph{H. sapiens} and Neanderthals, introgressed ancestry deserts
                should overlap with each other more often than expected by chance
        \end{itemize}
    \item Some conjectures about hybrid zones
\end{itemize}

\section*{Methods}

\subsection*{Computing expected genetic variance using the diffusion approximation}

\subsection*{Simulations with free recombination}

\subsection*{Individual-based simulations with linkage}

@@

\begin{figure}[tb!]
    \centering
    %\includegraphics{}
    \caption{
        \textbf{Dynamics of additive genetic variance in two populations with admixture.}
        Coupled with the equations for the decay of $V_G$ due to drift and selection,
        and the increase in $V_G$ through mutation, show that the expected dynamics
        are matched by simulations without linkage.
    }
    \label{fig:VG_dynamics}
\end{figure}

@@
